---
title: "r2Winsteps: An R package for interfacing between R and the Rasch Modeling Software Winsteps"
author: "Daniel Anderson"
date: "`r Sys.Date()`"
bibliography: r2Winsteps.bib
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The *r2Winsteps* package was developed to provide a convenient interface
  between *R* and the Rasch modeling software *Winsteps*. The package is not
  intended to encompass the full capabilities of *Winsteps*, but rather to
  provide a simple framework for estimating many commonly applied models. The
  primary features of the package include:

* Write control and data files for Winsteps with the `r2Winsteps()` function,
  which includes automatic detection of polytomous item types. Either the
  rating scale [default; @Andrich] or partial credit models [@Masters] can be
  estimated.
* Run Winsteps directly from R with the `runWinsteps()` function, which writes
  and executes a `.bat` file to call *Winsteps*. Item, person, and structure
  (if a polytmous model is estimated) parameters are returned in a list.
  Intermediary files (control, data, output, etc.) can be stored or discarded
  (default).
* Batch run a set of models with the `batchRunWinsteps()` files. Essentially
  calls `runWinsteps()`, but takes as its argument a list of
  data frames, with a different model fit and parameter estimates returned for
  each data frame in the list.

There were three primary motivation for developing *r2Winsteps*. First, 
  *Winsteps* is a powerful software that can fit a range of Rasch models with impressive speed. By linking the software to *R*, and unified framework is provided where users can prep, explore, and analyze their data within the same environment. This is important not only for convenience and efficiency purposes, but also to facilitate transparency. When multiple software packages are used, and multiple script files stored, it can be easy for an outside researcher (or indeed, the same researcher at a later time) to get lost in the process used. This is also an important principle of reproducible research [@Stodden], and can facilitate dynamic document generation [@Xie] with Winsteps, where all tables and figures are generated in a report are generated dynamically. That is, as the model or data are updated, so too are the features within the document that depend on them. This document is an example of using *Winsteps* within *R* to generate a reproducible and dynamic document, via the *r2Winsteps* package (although in practice, most documents would include the code being hidden).

The second primary motivation for developing *r2Winsteps* was the powerful
  plotting capabilities in *R*, coupled with tremendous flexibility. *Winsteps*
  provides many excellent plots, available through the graphs menu, but they
  are limited in that the user has little control over their appearance (e.g.,
  line types, colors, etc.). By contrast, *R* provides almost limitless
  flexibility, and multiple packages are available for facilitating plot
  creation (e.g., ggplot2 [@Hadley], lattice [@sarker]). The built-in plotting
  function in the *r2Winsteps* package are still under active development.
  These were all created with the base graphics, and should be fully
  customizable.

The third and final primary motivation for developing *r2Winsteps* was to develop an efficient method for batch processing analyses with *Winsteps*. While a batch processing option is available through *Winsteps*, one still must write each control and data file for each analysis. The *r2Winsteps* package allows for an efficient means of batch writing control and data files for analysis, which can then be estimated through *Winsteps* batch mode.

Note that the package is still in active development, and this vignette will be
  updated over time accordingly. The purpose of this vignette is to provide a
  few illustrated examples of using the package in its current stage. If you use the package, please reference it using the code below. If you find any bugs, please email them to daniela@uoregon.edu, or log them at https://github.com/DJAnderson07/r2Winsteps/issues.

```{r, reference}
citation(package = "r2Winsteps")
```

# Installation
For the time being, the *r2Winsteps* package is housed exclusively on *github*.
  Installation is straightforward via the *devtools* package. If you don't have
  *devtools* installed, you will need to first run the following:

```{r, devtoolsInstall, eval = FALSE}
install.packages("devtools")
```

Then, you just need to load the *devtools* package and install *r2Winsteps* directly from github.

```{r, install, eval = FALSE}
library(devtools)
install_github("DJAnderson07/r2Winsteps")
```

Winsteps is exclusively designed for Windows, and if you are on a Windows
machine you should be good to go from here. However, if you'd like to use a
Mac, you can do so through *Wine* (see https://www.winehq.org), a free, open-
source "compatibility layer" for running windows programs on OS X. For help
getting Winsteps installed with *Wine*, see the *Using Wine with Winsteps and
the r2Winsteps Package on Macs* documentation.

**One last Note:** Older versions of Winsteps included different (less) output
from the item and person files. The *r2Winsteps* package was built with
*Winsteps* Version 3.90. If you have an earlier version of Winsteps, there are
workarounds you can use in the options of the `read.pfile()` and `read.ifile()`
functions, and you may want to consider modifying the source code for your
particular installation.

# Using the package: Example 1, LSAT data
Now that everything is installed, let's start with a simple example. We'll
begin by fitting a Rasch model with dichotomous, using the supplied LSAT data.

```{r, loadData}
library(r2Winsteps)
data(LSAT)
head(LSAT)
```
It's generally a good idea to inspect some preliminary data, so we can get an
idea of what to expect from the model, and whether the assumptions of the model
appear tenable. First, we'll estimate the proportion of respondents responding
correctly to the items. Because the items are dichotomous, this is just the
mean.

```{r, pValues}
apply(LSAT[ ,4:8], 2, mean)
```
So all items appear somewhat easy, but Item 1 is clearly the easiest while Item
3 is clearly the most difficult.

Next, we can compute the point-biserial correlation, by correlating the
response vector for each item with a vector of raw scores. We'll compute the
raw scores, and then compute the correlations.

```{r, ptBiserial}
raw <- rowSums(LSAT[ ,4:8])
round( sapply(LSAT[ ,4:8], function(i) cor(i, raw)), 2)
```
These are classical test theory indicators of *item discrimination*. The Rasch
model assumes essentially equivalent item discrimination, and so we're looking
to see if any items appear wildly different from the others. It's also worth
noting that item-fit statistics reported by *Winsteps*, such as the mean square
outfit, are produced by evaluating the differences from the *average* biserial
correlation [see @Wu]. These all appear reasonable, so let's go ahead and fit
the model. In this case, because the model and data are straightforward, we
simply need to call the `runWinsteps()` function, which requires the data be
split into a data frame of item responses and a data frame of person
demographics.

```{r, LSAT_mod1}
pars_LSAT <- runWinsteps(LSAT[ ,4:8], LSAT[ ,1:3])

itms <- LSAT[ ,4:8]
dems <- LSAT[ ,1:3]

str(pars_LSAT)
plot(pars_LSAT, "TIF/IIF")
pars_ordered$ItemParameters
batch.pfile()

# START HERE

d <- read.csv("/Users/Daniel/Desktop/Ordered.csv")
pars_ordered <- runWinsteps(d[ ,1:25], d[ ,26:ncol(d)], partialCredit = TRUE)
str(pars_ordered)
pars_ordered$StructureFiles
sd <- split(d, d$Sex)

sdItms <- lapply(sd, "[", 1:25)
sdDemos <- lapply(sd, "[", 26:ncol(d))
pars_orderedGroup <- batchRunWinsteps(sdItms, sdDemos, list("Males", "Females"))
str(pars_orderedGroup)

```

The function writes control and data files necessary to run a basic Rasch model
for the supplied data. Additionally, a *.bat* file is written to the same
directory, which is then executed to call *Winsteps* and run the analysis. By
default, the control file is written to have *Winsteps* write out the person
and item parameter estimates. These are then read back into R through other
functions in the package. Note that one of the things that's nice about this
approach, however, is that the person files are read back in with each
demographic variable separated into its own column, whereas the Winsteps files
contain all person identifying information in a single column.

We can now inspect the results by subsetting the list. For example, we can look at items by

```{r itemsLSAT}
itms_LSAT <- pars_LSAT$ItemParameters
itms_LSAT
```
In this case, the test was really small (only five items) and so we can pretty much view everything we need here. In other cases, the output may be too large, and so further subsetting can be helpful. For example, let's look at the person parameters

```{r personsLSAT}
pers_LSAT <- pars_LSAT$PersonParameters
head(pers_LSAT)
sd(pers_LSAT$Theta)
```

The first six rows are shown above out of the `r nrow(pers_LSAT)` respondents. The mean square outfit is often a useful indicator not only of item functioning, but also of person fit to the model expectations (with high values indicating unexpected responses). Let's calculate the percentage of persons with a mean square outfit between 0.7 and 1.3 (these are fairly arbitrary cutoffs).

```{r personOutfit}
nrow( subset(pers_LSAT, Outfit >= 0.7 & Outfit <= 1.3) ) / nrow(pers_LSAT)
```

In the above code, I've calculated the number of rows in the subset data, and divided the result by the number of rows in the full data, to come up with the proportion `r nrow( subset(pers_LSAT, Outfit >= 0.7 & Outfit <= 1.3) ) / nrow(pers_LSAT)`. 





The above example is really the heart of the package, and the following
examples will show how it can be extended a bit. But it's also worth mentioning
that each step can be carried out individually. While the package is intended
to cover a wide range of models, it is not intended to cover the full
functionality of *Winsteps*. If there is a model that one would like to fit,
which is not included in the package, it can still be useful as a way to write
the baseline control file and corresponding data file with the function
`r2Winsteps()`. For example, we can use the same call as before, only this time
using the `r2Winsteps` function, rather than `runWinsteps()`. This will write
the corresponding control and data files, which can be modified by hand to
obtain the specific model one would like to run.

```{r writeCntrlDta}
r2Winsteps(LSAT[ ,4:8], LSAT[ ,1:3])
```



\newpage

# References